Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-08 introduction.tex
Updates
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@971 ccdd05d5-56c1-461a-bf98-85f6876523b4

As a compact binary system loses energy to gravitational waves, its orbital
separation decays, leading to a run-away inspiral with the
gravitational-wave amplitude and frequency increasing until the system
eventually merges soon after the innermost stable circular orbit (\ISCO). If a
neutron star is involved it may become tidally disrupted near the merger.  This
disrupted matter can fuel a bright electromagnetic counterpart in the merger's aftermath~\cite{shibata:2007}.
The coalescence of compact binary systems consisting of neutron stars
(\textsc{ns}) and/or black holes (\textsc{bh}) is the most promising source of
gravitational radiation for Advanced \LIGO, Virgo, \GEO\, and
\LCGT~\cite{ALIGOWeb, AVirgoWeb, GEOWeb, LCGTWeb}.  Tens of binary coalescence
events are expected to be observed in the advanced detector era later this
decade~\cite{Abadie:2010p10836}. In the era of first-generation
detectors the gravitational-wave community
initiated a project to send alerts when potential gravitational-wave transients
were observed in order to trigger followup observations by optical telescopes.
The best demonstrated latencies were 30--60 minutes~\cite{HugheyGWPAW2011}, which was
an important achievement, but far too late to catch prompt electromagnetic emission.

@@

To this end, we have the ambition of reporting gravitational-wave candidates
not minutes \emph{after} the merger, as has already been accomplished, but
seconds \emph{before}.  By examining the signal to noise ratio (\SNR{}) stream
for threshold crossings before the gravitational-wave signal leaves the
detection band, it is possible to trade some \SNR{}
for latency.  Figure \ref{fig:earlywarning} shows projected early trigger rates
for \textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming the event rate predictions
in~\cite{Abadie:2010p10836}.
%
\begin{figure}
\begin{center}
\includegraphics{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning} Expected number of \textsc{ns}--\textsc{ns} sources that will
be detectable $t$ seconds before coalescence.  The heavy solid line is the most
likely yearly rate estimate $\dot N_{\mathrm{re}}$ during Advanced \LIGO.  The
light solid lines represent the confidence interval $[\dot N_{\mathrm{low}}, \dot 
N_{\mathrm{high}}]$ described in
\cite{Abadie:2010p10836}.  Advanced \LIGO\ noise projections are described in 
\cite{ALIGONoise}.  Assuming that an \SNR\ of 8 is sufficient for
detection and that we observe $\dot N_{\mathrm{re}} = 40$ events~yr$^{-1}$ with a
detector in the `zero detuning, high power' configuration, about 10~yr$^{-1}$ will be 
detectable within 10 seconds before merger and $\sim1$~yr$^{-1}$ will be detectable 
within 100 seconds of merger.  The `zero detuning, lower power' and `\textsc{bbh} optimized' configurations are still more conducive to advance detection
but at the cost of fewer total detectable events above an \SNR\ of 8.}
\end{center}
\end{figure}

>>>>

As a compact binary system loses energy to gravitational waves, its orbital
separation decays, leading to a run-away inspiral with the gravitational-wave
amplitude and frequency increasing until the system eventually merges soon
after the innermost stable circular orbit (\ISCO). If a neutron star is
involved it may become tidally disrupted near the merger.  This disrupted
matter can fuel a bright electromagnetic counterpart in the merger's
aftermath~\cite{shibata:2007}.

Gravitational-wave detectors are sensitive the to these gravitational-wave
signals.  Indeed, the coalescence of compact binary systems consisting of
neutron stars (\textsc{ns}) and/or black holes (\textsc{bh}) is the most
promising source of gravitational radiation for Advanced \LIGO, Virgo, \GEO\,
and \LCGT~\cite{ALIGOWeb, AVirgoWeb, GEOWeb, LCGTWeb}.  Tens of binary
coalescence events per year are expected to be observed in the advanced
detector era later this decade~\cite{Abadie:2010p10836}. In the era of
first-generation detectors the gravitational-wave community initiated a project
to send alerts when potential gravitational-wave transients were observed in
order to trigger followup observations by optical telescopes.  The best
demonstrated latencies were 30--60 minutes~\cite{HugheyGWPAW2011}, which was an
important achievement, but far too late to catch prompt electromagnetic
emission.

@@

To this end, we have the ambition of reporting gravitational-wave candidates
not minutes \emph{after} the merger, as has already been accomplished, but
seconds \emph{before}.  By examining the signal to noise ratio (\SNR{}) stream
for threshold crossings before the gravitational-wave signal leaves the
detection band, it is possible to trade some \SNR{} and sky localization
accuracy~\cite{Fairhurst2009} for latency.  Figure \ref{fig:earlywarning} shows
projected early trigger rates for \textsc{ns}--\textsc{ns} binaries in Advanced
\LIGO\ assuming the event rate predictions in~\cite{Abadie:2010p10836}.
%
\begin{figure}
\begin{center}
\includegraphics{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning} Expected number of \textsc{ns}--\textsc{ns} sources that will
be detectable $t$ seconds before coalescence.  The heavy solid line is the most
likely yearly rate estimate $\dot N_{\mathrm{re}}$ during Advanced \LIGO.  The
light solid lines represent the confidence interval $[\dot N_{\mathrm{low}}, \dot 
N_{\mathrm{high}}]$ described in
\cite{Abadie:2010p10836}.  Advanced \LIGO\ noise projections are described in 
\cite{ALIGONoise}.  Assuming that an \SNR\ of 8 is sufficient for
detection and that we observe $\dot N_{\mathrm{re}} = 40$ events~yr$^{-1}$ with a
detector in the `zero detuning, high power' configuration, about 10~yr$^{-1}$ will be 
detectable within 10 seconds before merger and $\sim1$~yr$^{-1}$ will be detectable 
within 100 seconds of merger.  The `zero detuning, lower power' and `\textsc{bbh} optimized' configurations are still more conducive to advance detection
but at the cost of fewer total detectable events above an \SNR\ of 8.}
\end{center}
\end{figure}

### 2011-06-08 method.tex
Updates
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@971 ccdd05d5-56c1-461a-bf98-85f6876523b4

The first step of our proposed method is to divide the templates into
time slices in a \TD{} analogue to the \FD{}
decomposition described in~\cite{Marion2004, Buskulic2010}.
A matched filter is constructed for each time
slice.  The outputs form an ensemble of partial \SNR{} streams.  By linearity,
these partial \SNR\ streams can be suitably time-delayed and summed to
reproduce the \SNR\ of the full template.  To wit,
%
\begin{equation}
\label{eq:time-slices}
h_{i}[k] = \sum_{s=0}^{S-1}
	\cases{%
		h_i^s[k] & for $t^s \leqslant k / f^0 < t^{s+1}$ \\
		0 & otherwise
	}
\end{equation}
\editorial{Drew: Should this be $h_{i}[k]$ or $\rho_{i}[k]$?}
%
for $S$ integers $\{f^0 t^s\}$ such that $0  = f^0 t^0 < f^0 t^1 < \cdots < f^0 t^S = N$.
We will show
in the next section that this, combined with the \SVD{},
is sufficient to enable a computationally efficient \TD{} search and
furthermore is an essential part of an \earlywarning\ detection scheme.

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of \SVD\ {\em
basis templates}, previously demonstrated
in~\cite{Cannon:2010p10398}.  Similarly, the time-sliced templates described
above can be approximated to arbitrary accuracy by expansion into a set of
\SVD\ \emph{basis templates}, $u_l^s[k]$, related to the original time
sliced templates through the \emph{reconstruction matrix},
$v_{il}^s\sigma_l^s$:
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
%
\editorial{Say a word or two about \SVD\ tolerance?}%
%
The parameter $\numsvdtmps$ is the number of basis templates that are kept in
the approximation.  This determines the \SVD\ tolerance, which affects the
\SNR\ loss due to the approximation.  The authors of \cite{Cannon:2010p10398}
showed that high accuracy could be achieved with far fewer basis templates than
templates in the original template bank.  We find that when combined with the
time-slice decomposition, the number of basis templates \numsvdtmps\ is much
smaller than the original number of templates \numtmps\ and improves on
\cite{Cannon:2010p10398} by nearly an order of magnitude.  In the next section
we describe how we form our early-warning detection statistic using the time-slice
decomposition and the \SVD.

>>>>

The first step of our proposed method is to divide the templates into time
slices in a \TD{} analogue to the \FD{} decomposition described
in~\cite{Marion2004, Buskulic2010}.  To wit,
%
\begin{equation}
\label{eq:time-slices}
h_{i}[k] = \sum_{s=0}^{S-1}
	\cases{%
		h_i^s[k] & for $t^s \leqslant k / f^0 < t^{s+1}$ \\
		0 & otherwise
	}
\end{equation}
%
for $S$ integers $\{f^0 t^s\}$ such that $0  = f^0 t^0 < f^0 t^1 < \cdots < f^0
t^S = N$.  A matched filter is constructed for each time slice.  The outputs
form an ensemble of partial \SNR{} streams.  By linearity, these partial \SNR\
streams can be suitably time-delayed and summed to reproduce the \SNR\ of the
full template.  We will show in the next section that this, combined with the
\SVD{}, is sufficient to enable a computationally efficient \TD{} search and
furthermore is an essential part of an \earlywarning\ detection scheme.

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of \SVD\ {\em
basis templates}, previously demonstrated
in~\cite{Cannon:2010p10398}.  Similarly, the time-sliced templates described
above can be approximated to arbitrary accuracy by expansion into a set of
\SVD\ \emph{basis templates}, $u_l^s[k]$, related to the original time
sliced templates through the \emph{reconstruction matrix},
$v_{il}^s\sigma_l^s$:
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
%
The parameter $\numsvdtmps$ is the number of basis templates that are kept in
the approximation.  This determines the \SVD\ tolerance, which affects the
\SNR\ loss due to the approximation.  The authors of \cite{Cannon:2010p10398}
showed that high accuracy could be achieved with far fewer basis templates than
templates in the original template bank.  We find that when combined with the
time-slice decomposition, the number of basis templates \numsvdtmps\ is much
smaller than the original number of templates \numtmps\ and improves on
\cite{Cannon:2010p10398} by nearly an order of magnitude.  In the next section
we describe how we form our early-warning detection statistic using the time-slice
decomposition and the \SVD.
